---
title: "CoolR Usage Vignette"
author: "Dovetail Genomics"
date: "9/17/2020"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{HiCcompare Usage Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```

# Introduction

`CoolR` provides a lightweight interface to `cool` and `mcool` files using a native HDF5 R interface. You will get same information as provided by using `cooler` program, but the outputs will be provided as R objects that can be used for downstream analysis. 


# How to use `CoolR`

## Installation

The best way to install `CoolR` is from Bioconductor.

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("CoolR")
library(CoolR)
```

## Getting Cooler files 

In order to use `CoolR`, you will need the Hi-C interaction matrix stored in either `cool` or `mcool` format. To get a sense of this package, you can start with using the cool file for Omni-C data provided here: (give paths to omnic contact matrix here). Once you have a `cool`/`mcool` files, you are ready do use this package!

## Extracting data from `cool` files

`Cool` files store the data at a single bin size. For our example, the `cool` file has stored the contact matrix at 1000 bp resolution. To read the contact matrix for the entire genome into a `GInteractions` object, you can run following command.

```{r, eval = FALSE}
contact_matrix_whole_genome <- read.cool("Omni-C_1kb.cool")
```
